---
title: "Associating human age-related genes to human transcription factors"
author: "Wen Lim"
date: "2025-05-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# one-off installation -------------------------------------------------
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c(
  "readxl",      # read .xlsx
  "readr",       # read .csv
  "tidyverse",   # dplyr, purrr, ggplot2 …
  "decoupleR",   # enrichment stats + helper fns
  "dorothea"     # curated TF-target regulons
))

# every session --------------------------------------------------------
library(readxl)
library(readr)
library(tidyverse)
library(decoupleR)      # brings get_resource() & run_ora()
library(dorothea)       # provides dorothea_hs regulons

```

## Load data

```{r load data}

## ---- GenAge (Age-related genes) list --------------------------------- 
#https://genomics.senescence.info/genes/index.html
genage <- read_csv("genage_human.csv", show_col_types = FALSE) |>
          mutate(Symbol = toupper(symbol)) |>
          distinct(Symbol)

## ---- Alzheimer's related genes list ---------------------------------  #https://maayanlab.cloud/Harmonizome/gene_set/alzheimer/GeneRIF+Biological+Term+Annotations
AD_genes <- read_excel("Alzheimers_genes.xlsx",      
                  sheet = 1) %>%                # default is first sheet
  mutate(Symbol = toupper(Symbol)) %>%            # gene → upper-case HGNC symbol
  distinct(Symbol)

## ---- Human TF list --------------------------------------------------- 
# https://humantfs.ccbr.utoronto.ca
tfdb <- read_excel("Human TFs DatabaseExtract_v_1.01.xlsx") |>
        filter(toupper(`Is TF?`) == "YES") |>
        mutate(Symbol = toupper(`HGNC symbol`)) |>
        distinct(Symbol)

```

## Quick check which transcription factor genes are themselves in GenAge

```{r Quick check}

tf_age_overlap <- intersect(tfdb$Symbol, genage$Symbol)
length(tf_age_overlap)          # usually ~50
tf_age_overlap                  # print the symbols if you like

```

## Build the TF-target network to test

```{r }

## pull high-confidence DoRothEA interactions (levels A & B) ------------
data(dorothea_hs)   # loads a tibble called dorothea_hs

# Map evidence letters → numeric “likelihood” weights (optional but useful)
conf2like <- c(A = 1.0, B = 0.75, C = 0.5, D = 0.25, E = 0.05)

#Build the regulon
#Note: A regulon = one TF plus the set of genes it directly regulates, using DoRothEA resource. Each TF regulon is treated as a gene set, exactly like a KEGG pathway or GO term
regulon <- dorothea_hs %>% 
  filter(confidence %in% c("A", "B")) %>%           # strongest evidence only
  transmute(
    source     = toupper(tf),                       # rename 'tf' → 'source'
    target     = toupper(target),                   # target gene
    mor        = mor,                               # mode of regulation (+1/-1)
    likelihood = conf2like[confidence]              # numeric edge weight
  ) %>% 
  filter(source %in% tfdb$Symbol)                   # keep only regulators in TF db

```

## Hypergeometric enrichment: “Is this TF’s regulon enriched for GenAge genes?”

```{r }

#Hyper-geometric test answers: Is my list of ageing genes unexpectedly over-represented among the genes that TF is known to regulate?"
#i.e., whether the list contains more ageing genes that random chance allows

#Test calculates probability of drawing the ageing gene when you sample the regulon genes, from a population of total genes in any regulon and total number of ageing genes in the universe
#Taking intersection of TF list with GenAge list only gives TFs whose own genes appear in GenAge
#Using regulons lets you uncover TFs that regulate ageing genes, even if the TF itself is not annotated as an ageing genes e.g., master regulators

## define the universe --------------------------------------------------
age_in_univ <- intersect(genage$Symbol, regulon$target) # ageing genes present

## build the 1-column ‘matrix’ just like before
gene_score  <- matrix(1,                                # all 1’s = “hit”
                      nrow     = length(age_in_univ),
                      dimnames = list(age_in_univ, "age"))

## run ORA - over-representaiton analysis with one-tailed Fisher test inside
# n_up     = number of “hits” (all ageing genes)
# n_bottom = number of “anti-hits” (0, because the list is unsigned)

ora_res <- run_ora(
  mat      = gene_score,
  net      = regulon,
  n_up     = nrow(gene_score),
  n_bottom = 0,
  minsize  = 0          # keep even tiny regulons
)


## Compute per-TF statistics we care about -------------------------------
reg_stats <- regulon %>%                           # start from the edge list
  group_by(source) %>%                             # one row per TF
  summarise(
    RegulonSize = n_distinct(target),              # total genes it regulates
    AgeingHits  = n_distinct(                      # how many are GenAge genes
                    target[target %in% age_in_univ]
                  ),
    .groups = "drop"
  )

## 7·2  Join those counts back to the ORA table and add FDR -------------------
tf_ranked <- ora_res %>% 
  rename(TF = source) %>%                          # nicer column name
  mutate(FDR = p.adjust(p_value, "BH")) %>%        # Benjamini–Hochberg
  left_join(reg_stats, by = c("TF" = "source")) %>%# add hit counts & sizes
  select(TF, p_value, FDR, AgeingHits, RegulonSize,
         everything()) %>%                         # keep any extra cols (NES, etc.)
  arrange(FDR)

## 7·3  Quick sanity check -----------------------------------------------------
head(tf_ranked, 10)   # should list FOXO, NFKB1, E2F … with non-NA counts

```

## Hypergeometric enrichment: “Is this TF’s regulon enriched for Alzheimer's genes?”

```{r }

## define the universe --------------------------------------------------
age_in_univ_AD <- intersect(AD_genes$Symbol, regulon$target) # ageing genes present

## build the 1-column ‘matrix’ just like before
gene_score_AD  <- matrix(1,                                # all 1’s = “hit”
                      nrow     = length(age_in_univ_AD),
                      dimnames = list(age_in_univ_AD, "age"))

## run ORA - over-representaiton analysis with one-tailed Fisher test inside
# n_up     = number of “hits” (all ageing genes)
# n_bottom = number of “anti-hits” (0, because the list is unsigned)

ora_res_AD <- run_ora(
  mat      = gene_score_AD,
  net      = regulon,
  n_up     = nrow(gene_score_AD),
  n_bottom = 0,
  minsize  = 0          # keep even tiny regulons
)


## Compute per-TF statistics we care about -------------------------------
reg_stats_AD <- regulon %>%                           # start from the edge list
  group_by(source) %>%                             # one row per TF
  summarise(
    RegulonSize = n_distinct(target),              # total genes it regulates
    ADHits  = n_distinct(                      # how many are Alzheimer's genes
                    target[target %in% age_in_univ_AD]
                  ),
    .groups = "drop"
  )

## 7·2  Join those counts back to the ORA table and add FDR -------------------
tf_ranked_AD <- ora_res_AD %>% 
  rename(TF = source) %>%                          # nicer column name
  mutate(FDR = p.adjust(p_value, "BH")) %>%        # Benjamini–Hochberg
  left_join(reg_stats_AD, by = c("TF" = "source")) %>%# add hit counts & sizes
  select(TF, p_value, FDR, ADHits, RegulonSize,
         everything()) %>%                         # keep any extra cols (NES, etc.)
  arrange(FDR)

## 7·3  Quick sanity check -----------------------------------------------------
head(tf_ranked_AD, 10)   # should list FOXO, NFKB1, E2F … with non-NA counts

```

## Visualise 15 strongest TFs at a glance for Ageing and Alzheimer's genes

```{r XX}

library(ggplot2)

tf_ranked %>% 
  slice_head(n = 15) %>%                        # top 15 rows
  mutate(TF = forcats::fct_reorder(TF, FDR, .desc = TRUE)) %>% 
  ggplot(aes(TF, -log10(FDR))) +
  geom_col() +
  coord_flip() +
  labs(title = "TF regulons enriched for GenAge genes",
       y = expression(-log[10] * " FDR"),
       x = NULL)

tf_ranked_AD %>% 
  slice_head(n = 15) %>%                        # top 15 rows
  mutate(TF = forcats::fct_reorder(TF, FDR, .desc = TRUE)) %>% 
  ggplot(aes(TF, -log10(FDR))) +
  geom_col() +
  coord_flip() +
  labs(title = "TF regulons enriched for Alzheimer's genes",
       y = expression(-log[10] * " FDR"),
       x = NULL)

```

## Export the significant edges for Cytoscape 

```{r XX}

sig_tfs <- tf_ranked %>% 
  filter(FDR < 0.05) %>% 
  pull(TF)

regulon %>% 
  filter(source %in% sig_tfs) %>% 
  write_tsv("TF_Ageing_Regulon_SigEdges.tsv")

# → In Cytoscape: import as edge table; map 'mor' to edge colour,
#   'likelihood' to edge width; run network layout of your choice.

#	In the preview grid choose:
#	“source” for Source Interaction
#	“target” for Target Interaction

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Miscellaneous notes

```{r XX}

#Oct 3/4, Sox2 are not in transcription factor set
# --> Oct 3/4 are not even in the human transcription factor dataset to start with?

#Klf4 does not have a significant p-value


```
